Tropical mathematics and the Lambda-calculus I

### 2023-07-31 CSL24/section5bis.tex
commit

In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.

In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which accounts for differences in behavior. 
A fundamental insight coming from this line of work is that, if one can somehow   \emph{bound} the number of times that a program may duplicate its input, the resulting program will be \emph{Lipschitz-continuous}:  
if $M$ may duplicate at most $L$ times, then an error $\epsilon$ between two inputs will result in an error less or equal to $L\cdot \epsilon$ in the corresponding outputs \cite{Reed2010, Gaboardi2017}.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.

@@

At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions? This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps. In other words, unbounded term application could be seen 
as a limit of \emph{more and more sensitive} operations.

>>>>

In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, that a small error in the input will produce a comparably small error in the output. 
In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which accounts for differences in behavior. 
A fundamental insight coming from this line of work is that, if one can somehow   \emph{bound} the number of times that a program may duplicate its input, the resulting program will be \emph{Lipschitz-continuous}:  
if $M$ may duplicate at most $L$ times, then an error $\epsilon$ between two inputs will result in an error less or equal to $L\cdot \epsilon$ in the corresponding outputs \cite{Reed2010, Gaboardi2017}.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$.

@@

At this point, as the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{weighted relational semantics} \cite{Manzo2013}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions? This way, the Taylor formula could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps. In other words, unbounded term application could be seen 
as a limit of \emph{more and more sensitive} operations.

### 2023-07-31 CSL24/sectionLaw.tex
commit

Now, it turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories. Conversely, complete $\Lawv$-categories are \emph{tensored}, that is, admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
%
%
%To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
%two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
%$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
%$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
%Conversely, 
%for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
%$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.
%
%This leads to the following:
%
%
%\begin{theorem}[cf.~\cite{Stubbe2006}]
%$\Mod$ and $\GMet$ are isomorphic categories.
%\end{theorem}

>>>>

Now, it turns out that the categories $\Mod$ and $\GMet$ are isomorphic \cite{Stubbe2006}. First,  any $\Lawv$-module $(M,\preceq, \star)$ is a $\Lawv$-category via
$M(x,y) = \inf\left\{ \epsilon \mid \epsilon \star x\geq y\right\}$. Moreover, a homomorphism of $\Lawv$-modules is a functor of the associated $\Lawv$-categories. Conversely, complete $\Lawv$-categories are \emph{tensored} (cf.~\cite{Stubbe2014}), and thus admit a $\Lawv$-module structure $\epsilon \otimes x$ (see the Appendix).
Moreover, a continuous functor between complete $\Lawv$-categories is the same as a homomorphism of the associated $\Lawv$-modules. 
%
%
%To conclude our correspondence between $\Lawv$-modules and complete $\Lawv$-categories, it remains to observe that the 
%two constructions leading from one structure to the other are one the inverse of the other: for any $\Lawv$-module $(M,\preceq,\star)$,
%$x\preceq_{M}y$ iff $M(y,x)=0$ iff $x\preceq y=0\star y$, and, from  
%$M(\epsilon \star x, y)= M(x,y)\dotdiv \epsilon$, we deduce $\epsilon\otimes x=\epsilon \star x$. 
%Conversely, 
%for any complete $\Lawv$-category $X$ and $x,y\in X$, one can check that 
%$X(y,x)=\inf\{ \epsilon \mid X(\epsilon\otimes y,x )=0\}$.
%
%This leads to the following:
%
%
%\begin{theorem}[cf.~\cite{Stubbe2006}]
%$\Mod$ and $\GMet$ are isomorphic categories.
%\end{theorem}

### 2023-07-31 CSL24/sectionTaylip.tex
commit

The key ingredient is the notion of Taylor expansion of $\lambda$-terms, that we now quickly recall. 
Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
Observe that in a resource term applications are always \emph{bounded}: they may use an exact number of copies of their input.

The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as $\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{ t \langle u_{1},\dots, u_{k}\rangle \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$. 
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all resource terms
 $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tLs. 
However, resource term need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot (y\cdot x)$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

@@

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
The equation above is a nothing more than a tropical version of the Taylor formula \eqref{eq:taylor}:
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
%,  i.e.~it coincides with the tropical %interpretation of the
%version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:

@@

\begin{theorem}
Let $\mathcal S$ be one of $\RM{PCF}^{\Lawv},\STLC,\BSTLC,\STDLC$. Let $\Gamma\vdash_{\mathcal S}M:A$ and $a\in \model{A}$.
\begin{enumerate}
\item For $\mathcal S=\BSTLC$, $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;

\item For $\mathcal S=\STDLC$, if $t$ is a resource term, then $\model{\Gamma\vdash_{\mathcal S}t:A}^{!}_{a}$ is Lipschitz;

@@

\end{theorem}

@@

We conclude our discussion by presenting an interesting application of the Taylor expansion in $\LREL_{!}$: as proved in the previous section, all tLs are locally Lipschitz; now, Theorem \ref{thm:taylor} can be used to compute approximations of the Lipschitz constants of an actual higher-order program.

\begin{corollary}
Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tLs $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
 $B_{\delta}(\model{N})$.
\end{corollary}
{\color{red}ADD PROOF SKETCH}

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tLs $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= N+2 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

@@

% Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n ! \model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in ! \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
%Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$. Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n ! {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.

>>>>

The key ingredient is the notion of Taylor expansion $\Te{M}$ of a $\lambda$-term $M$, which is a set of terms of the differential $\lambda$-calculus defined as follows:
%
%
%
% which is defined as follows: 
%
%that we now quickly recall. 
%
%
%Let $M\langle N_{1},\dots, N_{k}\rangle$ be an abbreviation for
%$\mathsf D^{k}[M, N_{1},\dots, N_{k}]0$. A \emph{resource $\lambda$-term} is any term belonging to the grammar $t::= x \mid \lambda x.t \mid t \langle t,\dots, t\rangle$.
%%
%
%The Taylor expansion of a $\lambda$-term is a set of resource terms $\Te{M}$ defined inductively as 
%
$\Te{x}=\{x\}$, $\Te{\lambda x.M}=\{\lambda x.t\mid t\in \Te{M}\}$ and 
$\Te{MN}=\{t\cdot \langle u_{1},\dots, u_{k}\rangle  \mid k\in \mathbb N, t\in \Te{M}, u_{i}\in \Te{N} \}$, where $t\cdot \langle u_{1},\dots, u_{k}\rangle$ is an abbreviation for $\mathsf D^{k}[t, u_{1},\dots, u_{k}] 0$. 
Observe that in the terms appearing in $\Te{M}$ all applications are always \emph{bounded}: they may use an exact number of copies of their input. 
Such terms are usually called \emph{resource $\lambda$-terms} \cite{Pagani2009, Manzo2012}.
Considering the term $M=zxx$ from Example \ref{ex:zxx}, all terms
 $t_{n,m}=z \,\langle x^{n}\rangle\,\langle x^{m}\rangle$, for $n,m\in \mathbb N$, are in $\Te{M}$. 
Observe that the interpretation of $t_{n,m}$ yields a tropical polynomial
$\model{t_{n,m}}^{!}(x)(z)= y_{[n,m]}+(n+m)x$, rather than a tps. 
However, resource $\lambda$-terms need not always yield a tropical polynomial. For instance, consider $y:(*\to*)\to (*\to*), x:(*\to *) \vdash t:  (*\to  *)$
with $t=y\cdot \langle y\cdot \langle x\rangle \rangle\in \Te{y(yx)}$. Then 
$\model{t}^{!}: \Lawv^{  \mathcal M_{\mathrm{fin}}(\mathbb N)\times \mathbb N}
\times \Lawv^{\mathbb N} \to \Lawv^{\mathbb N}$ is given by
$
\model{t}^{!}(y,x)_{i}= \inf_{m,n\in \mathbb N}\big\{    y_{[m],i}  +  y_{[n],m}+  x_{n}\}
$, 
which is not a polynomial. Yet, $\model{t}^{!}$ is Lipschitz, more precisely, 1-Lipschitz in $x$ and 2-Lipschitz in $y$. This is a general fact, as shown below.

@@

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{!X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
The equation above is a nothing more than a tropical version of the Taylor formula from the Introduction:
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$.
%,  i.e.~it coincides with the tropical %interpretation of the
%version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.

Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:

@@

\begin{theorem}\label{thm:taylor}
Let $\mathcal S$ be one of $\RM{PCF}^{\Lawv},\STLC,\BSTLC,\STDLC$. Let $\Gamma\vdash_{\mathcal S}M:A$ and $a\in \model{A}$.
\begin{enumerate}
\item For $\mathcal S=\BSTLC$, $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is a tropical polynomial, and thus Lipschitz;

\item For $\mathcal S=\STDLC$, if $M\in \Te{M'}$, then $\model{\Gamma\vdash_{\mathcal S}M:A}^{!}_{a}$ is Lipschitz;

@@

\end{theorem}
\begin{proof}
1). As we already observed, the interpretation of a bounded term is a tropical polynomial.
2.) From Proposition \ref{prop:troplinear} 2.~observing that a resource term $t(x)$ may use a variable $x$ a fixed number $n$ times, so that its matrix lies in $\Lawv^{!_{n}X\times Y}$. 
%Now we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and by taking the maximum Lipschitz constant among the finite number $\mathrm{Card}(\model A)$ of them, we obtain the thesis.
3). From Theorem~\ref{thmTLSlocLip}.
4). It follows from \autoref{cor:T(M)=M} plus the fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

@@

We conclude our discussion with an application of the Taylor expansion in $\LREL_{!}$: as proved in the previous section, all tps are locally Lipschitz; now, Theorem \ref{thm:taylor} can be used to compute approximations of the Lipschitz constants of an actual higher-order program.

\begin{corollary}
Suppose $x: A \vdash_{\STLC}M:B$ and $\vdash_{\STLC} N:A$. 
Then for all $t\in \Te{M}$ such that $\model{t}^{!}(\model{N})\neq \infty$, and $\delta>0$, the tps $\model{ x:A\vdash_{\STLC}M:B}^{!}$ is $\frac{\model{t}^{!}(\model{N}+2\delta)}{\delta}$-Lipschitz over the open ball
 $B_{\delta}(\model{N})$.
\end{corollary}
\begin{proof}
We apply the Lipschitz estimate from Thm.~\ref{thmTLSlocLip} for $\model{M}^{!}$,  and use the fact that, from Thm.~\ref{thm:taylor} 4.~it follows that $\model{t}^{!}\geq \model{M}^{!}$.
To conclude, since $\model{t}^{!}$ is concave and non-decreasing, $\max_{\overline{B_{3\delta}(x)}}\model{t}^{!}= \model{t}^{!}(x+3\delta)$, which yields the stated Lipschitz estimate. 
\end{proof}
%{\color{red}ADD PROOF SKETCH}

\begin{example}
Consider again the term $M=zxx$ from Example \ref{ex:zxx}-
%, satisfying $x:*, z:(*\to*\to*) \vdash M:  *$. Since $\mathcal M_{\mathrm{fin}}(\{*\})\simeq \mathbb N$, we have that 
%
% $\model{M}^{!}: \Lawv \times \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}\to \Lawv$ is given by
 The (generalized) tps $\model{M}^{!}(x)(y)= \inf_{n,n'\in \mathbb N}\{y_{[(n,n')]}+(n+n')x\}$ is not (globally) Lipschitz: for any 
$L>0$, choose a natural number $N>L$, let $Y\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ be such that $Y_{\mu}<\infty$ only if $\mu=[(n,n')]$ with $n+n'\geq N$; then $|\model{M}^{!}(x)(Y)- \model{M}^{!}(x+\epsilon)(Y)|\geq N\epsilon > L\epsilon$. 
Now take the approximant $t= z \langle x^{N-1}\rangle \langle x\rangle \in \Te{M}$ (chosen so that $\model t^{!}(x)(Y)<\infty$). Its interpretation is the monomial 
$\model{t}^{!}(x)(Y) = Y_{[(N-1,1)]}+Nx$. We can then compute a Lipschitz-constant for $\model{M}^{!}$ around $\langle x,Y\rangle$
as $\frac{1}{\delta}\model{t}^{!}(\langle x,Y\rangle+\delta)= 3N+3 + \frac{Y_{[(N-1,1)]}+Nx}{\delta}$. 
\end{example}
%    
%    %
%  Consider now $F\in \Lawv^{\mathcal M_{\mathrm{fin}}(\mathbb N\times \mathbb N)}$ given by $F_{\mu}=0$ if $\mu=[(1,1)]$ and $F_{\mu}=\infty$ otherwise. 
%  Then $\model{M}^{!}(\langle x,F\rangle)=2x$
%%We show that we can compute a local 
%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

@@

% Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n ! \model {A_i}\times \model B}$, the tps $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in ! \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
%Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$. Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n ! {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
